HMMotion: Using tracking data to predict coverage strategies in American Football

### 2024-12-30 submission.Rmd
.

## Introduction
Deciding between man and zone coverage is one of the most critical strategic choices a defensive coordinator must make before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns. These predictions, however, seem to be, from a viewership's assessment, limited to specific plays without incorporating pre-snap motion which is a crucial element of modern offensive strategies. `r colorize("Ich weiß nicht, ob ich das so formulieren würde... Wir wissen nicht was deren Modell alles inkludiert. Es kann gut sein, dass sie auch tracking und daher auch premotion Daten nehmen. Oder glaubst Du ist das eine Expertise von einer Person mit domain knowledge und kein statistisches Modell? Rouven: Ne glaube das ist ein stat. Modell aber habe mir ein paar Szenen angeschaut und da kam die Prediction eigentlich immer nur wenn sich keiner bewegte.","red")`

Our project takes this model a step further. While we similarly predict man or zone coverage when the teams are set, we further leverage the additional information of pre-snap player movements. Using a hidden Markov model (HMM), we model defenders' trajectories based on hidden states, which represent the offensive players they may be guarding. This probabilistic approach allows us to infer dynamic coverage patterns during plays with pre-snap motion. To exploit this high-dimensional time-series data, we calculate the entropy of the state probabilities, thus, producing measures of uncertainty for each play. Incorporating these as covariates into the existing pre-motion model significantly improves both the AUC and detection accuracy.

In doing so, we present a data-driven framework that evaluates the effectiveness of pre-snap motion in uncovering defensive strategies, providing real-time tactical insights for coaches.

@@

Within these plays, we concentrate on the tracking data after the line has been set (because we are not interested in how players come out of the huddle) and before the ball has been snapped by the Center.

@@

\textbf{UMSCHREIBEN} \textit{For all players and the ball carrier we use the features already contained in the tracking data, namely x- and y-coordinates, speed, acceleration, distance covered, orientation and direction. Except for the ball carrier we further compute the euclidean distance, x-distance and y-distance to the ball carrier. For defensive players only, we additionally compute the absolute difference of the defender’s direction and the angle of the shortest segment between the defender and the ball carrier. Subsequently, we order all players (in each frame) with respect to their euclidean distance to the ball carrier and standardize all features.
More details regarding data pre-processing can be found in the Appendix.}

@@

Using t

#### 3. Post-motion prediction
We re-train the pre-motion model to predict whether the defense plays a man- or zone coverage scheme, however, in this step, we incorporate results from the HMM analysis as further covariates. In particular, we include the state-switching probabilities....

>>>>

## Introduction
Deciding between man and zone coverage is one of the most critical strategic choices a defensive coordinator must make before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers coverage predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns). However, these models seem to be trained on plays without pre-snap motion, or at least to the situations before motions (see [Amazon](https://aws.amazon.com/de/blogs/machine-learning/identifying-defense-coverage-schemes-in-nfls-next-gen-stats/)), which is a crucial element of modern offensive strategies.

Our project takes this model a step further. While we similarly predict man- or zone coverage when the teams are set, we further leverage the additional information of pre-snap player movements. Using a hidden Markov model (HMM), we model defenders' trajectories based on hidden states, which represent the offensive players they may be guarding. Incorporating summary statistics of the probabilistic HMM results as covariates into the existing pre-motion model significantly improves both the AUC and detection accuracy and further allows for evaluating the effectiveness of pre-snap motion in uncovering defensive strategies, providing real-time tactical insights for coaches.

@@

@@

@@

This probabilistic approach allows us to infer dynamic coverage patterns during plays with pre-snap motion. To exploit this high-dimensional time-series data, we calculate the entropy of the state probabilities, thus, producing measures of uncertainty for each play.

#### 3. Post-motion prediction
We re-train the pre-motion model to predict whether the defense plays a man- or zone coverage scheme, however, in this step, we incorporate results from the HMM analysis as further covariates. In particular, we use the aforementioned state probabilities of guarding the different offensive players in each second.
